% !TEX root = ../main.tex

\section{Introductory Remarks}

\input{figures/examples.tex}

%\textblue{Opening sentence.}

% Stems from ChatGPT:
%What if the technologies we [use] are quietly reshaping the boundaries of privacy and control?"
% innovations that... betray us
% When privacy is reduced to a setting and security to a feature...
% darker side of convenience
% risks that are normalized and overlooked

%Security is not just about keeping threats out; it is about questioning who we let in, and why. When privacy is a `setting' and security is a `feature,' our research community should continue looking at new ways of promoting insight and awareness over the shifting boundaries of privacy and control. 

Critical design is commonly attributed to Dunne and Raby~\cite{DuRa01,Dun05} however it can be seen as a refinement of earlier movements in arts and design, beginning with the Italian radical design movement and extending through critical practice in HCI~\cite{Mal17}. Consider the three examples provided in Table~\ref{tab:examples}. Each describes the design of an object or technology. In contrast to industrial design, these technologies are not oriented around problem-solving and the creation of solutions that are functional, useful, and pleasurable---they are not meant for mass production. Rather they offer a subtle criticism by illustrating something---electromagnetic radiation, the inevitability of death, the ubiquity and stealth of surveillance---that typically goes unquestioned. They are designed to invoke a dilemma for the user, raising questions, challenging values, and stimulating debate. Critical design considers these goals to be the function (or `para-function'~\cite{Dun05}) of the concept. %, enabling traditional design approaches to maximizing this functionality. 

Critical design has significant overlap with conceptual art, utilizing common methods, and has been disseminated through exhibitions and the art literature. However critical design works by showcasing things that are plausible, believable, and easy for the reader to situate, requiring design practice. Viewing critical design as art creates challenges. Consider the compass table (from Table~\ref{tab:examples}) in the context of an art gallery exhibit---it is meant to illustrate electromagnatic radition in every day life, in one's home, in different positions and orientations within that home---all of this is lost when it sits statically in a gallery. Should critical design be displayed in galleries, a vignette or film or narrative will often accompany the object to describe its context of use.

%\paragraph{Relation to Science.} 
%
%Critical design is ideological. Science can be as well. Even cryptography, positioned as one of the most scientifically rigorous pursuits in the field of security and privacy, is ideological, as argued eloquently by Rogaway in a recent keynote~\cite{Rog15}. For example, he notes that work on bland-sounding subjects like differential privacy and identity-based encryption are premised on a certain values and world views---respectively, one of massive data collection and one of government-escrowed master decryption keys. His point is that work in these areas \textit{affirm} certain ideologies; it is \textit{affirmative} design and more cryptographers should consider what is it that their designs affirm?
%
%From a different angle, Herley and van Oorschot question which aspects of security, privacy, and cryptography are even a proper science~\cite{HvO17}. Critical design is certainly not an objective form of design where success can be measured. We purposely do not set out to `measure' the success of our designs as this deeply misses the point of critical design. While our designs are for scientists, engineers, and technologists, it is not to establish theorems or facts but to raise new questions and communicate risks and invasions in a new way.

\subsection{Contributions}

\paragraph{Security \& Privacy.} Security and privacy risks are often intangible and invisible, making concepts difficult to communicate to everyday users. We saw a parallel to electromagnetic radiation in this regard, and considered if critical design could similarly aid in illuminating risks. The primary contribution of this paper is a set of designs.

\paragraph{Vignettes.} Critical designs are commonly described through vignettes. Occasionally they are built and displayed. In the case of Dunne and Raby's Compass Table (and the seven other objects in their \textit{Placebo} project~\cite{DuRa01}), the table was built and given to individuals to use. This field study was followed by an exit survey on how the participants felt about the object. In this paper, we describe our designs through short vignettes that illustrate its design and use case. If they are a successful design, the reader should easily imagine its use in her life and question it critically, resulting in its `rhetorical use'~\cite{Mal17} rather than practical use. These designs are transgressive and intended to provoke the reader's emotions; invoking humour, satire, ridicule, poetry, playfulness, lewdness, appropriation, irreverence or deliberate overcompensation.

\paragraph{Methodology.} In order to generate designs, we initially employed a XXX methodology, using concepts and sticky notes (see Figure). The design process coincided with the emergence of machine learning tools, based on large language models, like ChatGPT. We decided to also experiment with AI-assistance. We conducted four experiments. First, we (a) used traditional design methodology (no AI) to generate 10 designs (Section~\ref{}). We then (b) fed the 10 designs to ChatGPT version GPT-4g and asked it to generate similar designs (Section~\ref{}). It effortlessly generated designs that initially seemed sound, but under greater scrutiny, were lacking. We then (c) experimented with giving ChatGPT a sketch of what a possible design could involve and then dialogued with it to develop and refine the design, using it is a collaborative tool (Section~\ref{}). We found this approach produced good results and was less effort than method (a). Finally we (d) educated ChatGPT on the entire design process (Section~\ref{}), trying to involve it in the earliest stages of design (as opposed to (c) where we involved it after we had an initial sketch of an idea). We found this approach was less effective. In our experience, ChatGPT is capable of assisting in designs but requires human guidance and intervention. AI's strength is in generating large volumes of ideas which can be curated and refined by humans.

\subsection{Related Work}

\paragraph{Security and Privacy in Critical Design}

Much has been written on critical design and Malpass provides an excellent survey~\cite{Mal17}. Since our focus is security and privacy, we only summarize the following concepts, situated within or close to critical design, since they deal with issues of privacy: 

\begin{compactlist}
\item \textbf{The Pillow.} Dunne and Gaver's design takes the form factor of a transparent pillow and plays radio signals it picks up from the technology of that time: \eg phones, pagers, and baby monitors~\cite{DuGa97}. It challenges users to consider how confidential these signals are.  
\item \textbf{Open Informant.} Described in Table~\ref{tab:examples}, this design exposes nation-state surveillance.
\item \textbf{United Mirco-Kingdoms.} Dunne and Raby reimagine the UK as four `supershires' where one is occupied by `digitarians:' extreme technocrats ruled by algorithms in a state of total surveillance.  
\item \textbf{Alternet.} While not critical design, Gold's Alternet has the similar intention of provoking thought through design~\cite{Gol14}. The internet is re-imagined with users retaining full control over their private data, touching on web tracking and our growing digital dossiers. 
\end{compactlist}

We are unaware of any attempt to build a systematic set of examples based on security or privacy, nor much attention to it from academic research papers. %Many critical designs are `one-off.' The idea of developing a set of concepts related by a common theme, rather than individual concepts, is however established. For example, the Placebo Project mentioned already offered 8 concepts  (including Compass Table in Table~\ref{tab:examples}) that were each related to EMF radiation. 

%  Some security and privacy research offers seemingly unintentional critical design. In this case, we subsume the designs in our examples while clearly noting when they originate from other researchers. We also include `found' examples of critical design from outside of academia. We believe that situating them alongside our original ideas allows the reader to consider how close our concepts are to reality. 

\paragraph{Security and Privacy in Alternative Pedagogy}

But there are works on finding alternative ways (like design fiction~\cite{loureiro-koechlin_vision_2022}, role playing game~\cite{merrill_security_2020}, science fiction and design fiction~\cite{wong_real-fictional_2017}) to explain tricky concepts not only in security but also in other fields like sensing technologies.


\paragraph{Critical Design in HCI}

Critical design and speculative design are used in HCI literature~\cite{bardzell_what_2013}. 

To analyse and question the usage of a certain technology and the consequences of it~\cite{lawson_problematising_2015}

\paragraph{Methodologies for Design Creation} 
While design process can be based on well-established methods like whiteboarding, sticky notes, xxx; the integration of AI to the creation process has been proliferated with the abundance and the variety of AI-based tools. These tools can be used to generate the content itself or help the designer along the process of creation and ideation. There works that examine how users prompt generative AI (~\cite{sanchez_examining_2023}, ~\cite{chang_prompt_2023}); utilize generative AI for co-ideation or co-creation~\cite{tholander_design_2023},\cite{chiou_designing_2023}.

There are also efforts towards creation of computational tools to that would assist task that are manually done in general. One line of work is about creating tools to group design aspects to help ideation and exploration of design solutions\cite{huang_designnet_2023}. Speculative design is also used to imagine the future of the tools based on generative AI~\cite{lin_generative_2023}. Custom-made AI tools are proposed for specific design goals like integrating context awareness\cite{fan_contextcam_2024}.

\paragraph{Methodologies for Design Evaluation} 
While we only created the designs and did not apply any techniques to evaluate them, there are several methodologies for this purpose. User studies like interviews and surveys are one of the most common techniques employed~\cite{sanchez_examining_2023}, ~\cite{chang_prompt_2023}. They are deployed to evaluate different aspects of the design process like assesing the utility of the AI-tools used during the design process, understanding users' needs or deploying a participatory workshop. Methods to asses the creativity of AI-generated content are also proposed~\cite{chakrabarty_art_2024}.

\paragraph{Methodologies for Pedagodgy} 
CS education literature has explored how to teach concepts related to computer science in general or cybersecurity to a diverse audience. Some of which are understanding how CS students perceive AI and cybersecurity~\cite{ojha_computing_2023}, using visual cryptography for explaining concepts toK-12 students~\cite{rayavaram_designing_2023}.


% = = = = = = = =



\section{Methodology 1: Traditional Design}

\subsection{Methodology}

Our research goal was to create critical design vignettes illustrating security and privacy issues. We will first describe our design methodology for human designers. In later sections, we will explore the use of generative artificial intelligence. We began by reviewing the critical design literature, and we settled on delivering vignettes, which would be short paragraphs describing a new technology and illustrating, with short narratives, how it functions and the feelings it might invoke. The technology is meant to be plausible in the near future (the reader can conceive why someone, somewhere might think it is a good idea), while also showing the darker side of convenience, risks that are normalized and overlooked, or ways a user might be betrayed by the technology. During this period, we were encouraged to notice `found examples' of critical design in the wild, although perhaps not expressed that way.\footnote{For example, well-known cryptographer Matthew Green tweeted: \textit{There’s an entire class of browsing many folks will feel uncomfortable doing if their browser has a little icon in the window that says “Hi <User> I know this is you!” This is Human Behavior 101}.}

To generate designs, after some trial and error, we settled on a digital (Apple Freeform) `sticky note' approach, where we developed ideas within four groups: (1) security and privacy concerns, misconceptions, or principles that could be the basis of a critique; (2) domains where users might especially care about security or privacy; (3) a form factor for the new technology (\eg ); and (4) a possible user response (\eg ). Figure~\ref{} shows our board.\footnote{Artefacts such as our board available on GitHub repository (linked removed for peer-review anonymity)}. We brainstormed around individual notes and drew random combinations to spur discussion and thought.

The design methodology provided structure but did not make designing simple. Each of our 10 designs was quite difficult to develop, each taking many hours of discussion and dozens of discarded ideas that we could not translate into designs. 

\subsection{Designs}

\design{Crystal Avatars}{Crystal avatars are created instantly but only gets better over time. At first, they are not recognizable and cannot be differentiated from other users, but as users browse the site, small details crystallize. Alice is happy to save time registering for the site and have a personal avatar, which automatically added a cat accessory after she spent a lunch hour reading about hairball treatments. However she grows uneasy about the "Woman, Life, Freedom" badge, a cause she believes in but only vocalizes discriminately.}

\design{Calm Watch}{Calm Watch is a smart watch that tracks anxiety levels based on several biomarkers. When the anxiety level of the wearer crosses a threshold, a haptic vibration pulse prompts the user to look at the watch. From there, they can select a variety of calming techniques, including breathing exercises and guided meditation. Bob is nervous to talk to Alice and becomes flustered when his watch starts vibrating mid-conversation, just loud enough for Alice to hear it and notice his fidgeting with the watch.}

\design{Chatty}{The latest software update for Alice's voice-activated home assistant Chatty adds machine learning to better infer Alice's commands, such as adding items to her shopping list and suggesting music she might like. It does not always know whether Alice is talking to it or not, so over time it picks up on pieces of Alice's life and has trouble unlearning them. Alice is startled but pleased when Chatty chimes in with the name of `that actor who was in that thing' she was telling her friend about. She is bemused to find a specialized vinyl cleaning solution on her shopping list after she said Bob's records smells fishy.}

\design{Dynamic Laughtrack}{Digital television content, such as sitcoms, encode laughtracks as a series of cues rather than an actual recording. Smart TVs listen and classify the viewer's level of laughter on a 10 point scale (with fine grain training over time). The TV adds laughter to the content in a sensitive and considerate manner, where the laughtrack is only marginally higher on the scale than the viewer's own laughter. This gently nudges the viewer to greater enjoyment of the program without bombarding her. It also mixes in actual past recordings of the viewer's own laughter to capitalize on emotional mirroring.}

\design{Eleventh Finger}{Biometric authentication based on fingerprints is generally user-friendly and fast. Eleventh finger is a 3D printed rubber finger, customized with the user's fingerprint. You can put it on a keychain and use it on cold winter days when you do not want to remove your gloves. Alice lets her friend Bob borrow it when he stays at her house. It can also serve as a backup if the worst happens.}

\design{GOTTCHA}{GOTTCHA is a human-detection system to protect online accounts being accessed by bots. It invokes the device camera to analyze if a live human is using the device. To protect against video replays or machine-generated video, it unexpectedly prompts the user with a randomly selected image and carefully measures their reaction to it. Micro-expressions are involuntary facial displays of emotion that are too fast to mimic artificially. GOTTCHA's image bank can provoke disgust, anger, fear, sadness, happiness, surprise or contempt. Over time, Alice stops visiting websites that use GOTTCHA because of its capricious tendency to mix in disturbing images.}

\design{No(i)sy elevator}{Elevator music has never been regarded as elevating your mood. This is why employees will be happier riding the noisy elevator. When it recognizes a user, it spins up a selection from their most-played songs on major music streaming platforms. Alice finds hearing her favourite Miles Davis song grounds her at the start of each work day, while Bob and Carol, coincidently riding the elevator together, discover their shared love for 90s britpop. David is a bit more concerned about the profanity ridden banger that the nosy elevator plays for him and his boss.}

\design{Receiver plant}{The receiver plant in the front yard of Alice records the information of the people that pass by. It is a greedy plant that needs to absorb enough bluetooth data to grow. Bob is singlehandedly responsible for its lustrous canopy from walking by Alice's house every day, mostly while checking his work emails on his phone. Perhaps the plant overstepped when it displayed, "you are 15 minutes late today, are you sure you can make it to work on time?"}

\design{ToS Fishing}{ToS Fishing is a digital game where players earn points by hunting down excessive legalese on websites, including terms of service (ToS), privacy policies, and cookie policies. To play, users simply copy and paste the URL of the legalese. If the ToS has been seen before, the users are awarded points based on its word count. If it has not been seen before, it is reviewed by a human for validity and word count—and in this case, the user gets a finder's bonus. To mitigate people from launching custom websites, the game only accepts sites from the Alexis top million. Users display their aggregate score on a leaderboard with a profile showing their top catches (word count only, not the website), and can earn various badges for playing consistently.}

\design{Wifi Projector}{Minimalist wifi routers have a single light indicator to show internet connectivity, such as a green light as opposed to red light. This understates the interesting and intriguing data flowing through the device as users browse the internet. By contrast, the wifi projector is a talkative router that projects visualizations onto the ceiling above where it is placed. Every website that pulls in scripts and cookies from other domains is displayed as a constellation of stars of various colours and sizes, which slowly fade away. While Alice's visit to dictionary.com is a beautiful galaxy, she is also perturbed by the number of ads and tracking companies surveilling her.}

\subsection{Discussion}

\begin{itemize}
\item Example: Eleventh Finger. Biometrics (facial recognition), Security not privacy, STRIDE gives us spoofing (authentication), we were critical 2fa, funny (dark), biometrics is a shiny, critical is not as secure as it seems, usability principles (accessibility, losing), miss it from iphones, disney in the news, where (phone, home), 
\item Methodology was difficult (ratio of notes to designs)
\item Hit on why it is hard: methodology sets the stage but can't push out designs
\end{itemize}


\section{Methodology 2: LLM final}

\subsection{Methodology}


\begin{itemize}
\item We started before ChatGPT, it came out, vignettes are the kind of thing that LLMs should be good at.
\item We thought we would experiment with it
\item We prompted with a bit about critical design, we showed it our 10 designs, asked for more. Transcripts.
\end{itemize}

\subsection{Designs}

A good one:

\design{Transparent Wallet}{Alice's digital wallet automatically categorizes her purchases and shares anonymized spending habits on a public leaderboard. She enjoys seeing her eco-friendly purchases earning her top marks in the green consumer category. However, Alice is troubled when she notices ads for luxury items appearing more frequently after her high spending months are tracked and shared by the wallet.}



\subsection{Discussion}
\begin{itemize}
\item Transparent wallet
\item What we like: gamification is something we talked about, reasonably plausible, similar style
\item What we don't like: not clear, we didn't `get it' right away, name is just ok
\item We asked for 5 designs: they were all as good (link to transcript). For space, don't include
\item Repetitious: three designs on data sharing (weaponized against the user, twice in the form of ads and once in the form of insurance premiums). 
\item Not designs: hiring bot that is biased or algorithm for personalized social media feed: valid criticism, but have already been made (AI ethics 101), exist not speculative
\item Titles are boring
\item We tried for subtle, humorous, variety - not reflected

\end{itemize}

\section{Methodology 3: Polish}

\subsection{Methodology}

\begin{itemize}
\item We were not satisfied
\item The potential we saw from LLM was it can generate lots of ideas
\item The deficit was the ideas needing intervention from us
\item The idea of method 3 was to draw on all strengths: LLM can generate lots of idea, we can curate, iterate, inject our own ideas
\item Prompted with same prompt as method 3 for background (background + our designs), we gave rough notes of an "interesting idea" and then iterated with it to develop the idea.
\end{itemize}


\subsection{Designs}

\design{BloodKey}{BloodKey is a form of two-factor authentication that can protect online accounts from stolen passwords. With BloodKey, after providing their password, users authenticate with a pin-prick blood test using an inexpensive biomedical Bluetooth gadget. BloodKey was quickly adopted by leading websites because it was very hard to fake, despite its invasive nature. However, Alice grew tired of the daily inconvenience of pricking her finger, so she gradually stopped logging into her social media accounts altogether, while Bob's medical condition prevented him from using it at all. A media frenzy ensued when the most popular free-to-play online game sold users' genomic data to pharmaceutical companies, boosting their revenue streams.}

\design{Civic Compass}{Civic Compass is an add-on for popular GPS navigation apps that highlights the ethical considerations of various routes, such as avoiding surveillance hotspots or environmentally protected habitats. Public policy debates have emerged over the app's potential abuse by criminals. When Alice is in a rush, she ignores its suggestions. She is also disturbed to see that most of the surveillance hotspots are in impoverished, marginalized neighborhoods.}

\design{ElephantMind}{ElephantMind is an augmented reality (AR) social interaction app designed to enhance small talk by reminding users of past conversations. Alice cancels her coffee date with Bob when ElephantMind's service is down, fearing she won't remember their last chat. When Alice enthusiastically asks John about his trip to Paris, not knowing he was there for a difficult family matter, it leads to an awkward moment. Bob feels their friendship weakening as ElephantMind constantly pushes him to talk about trending political issues with Alice.}

\design{Profiled Plates} {Profiled Plates is a technology for high-end restaurants that offers an extraordinarily personalized dining experience by using extensive data mining and online profiling. With just a reservation name, the restaurant gathers detailed information about guests and their friends from social media, public records, and online interactions to create a bespoke meal tailored to individual preferences and backgrounds. Alice books a table for her birthday and is thrilled to find her favorite flowers on the table and a menu that excludes cilantro, a dislike she once tweeted about. However the system led to a low tip from Bob when it served a meal highly customized for a different Bob Bradley.}

\design{Square One}{Square One is a smart deletion feature for social media accounts which resets interests, deletes content, and unfollows people based on a simple prompt from the user about what they want removed. Jack uses Square One to unremember a volunteer project he led, only to find that potential employers now see him as less community-minded and pass him over for jobs. Jack is dismayed as he watches these projects falter under less-qualified hires.}

\subsection{Discussion}



\design{Rough Notes (ElephantMind)}{Avoiding small talk; to know more about the person you meet; the app reminds you the last conversation you had with them; constantly reminding something trivial (Alice promised Bob to grab a coffee sometime. Next time Bob sees Alice the app reminds Bob what Alice said.); Summarize the last conversation you had with them; For a person you meet for the first time, device gives you some background info about them so that you choice the right topic of conversation.; Name for the app: something that is about never forgetting like having a memory like an elephant; Photographic memory; Or something related to recording everything}

\begin{itemize}
\item ElephantMind
\item Prompt for names, might say: Memori, ElephantMind, Recall, Memento,  MindVault. Choose or iterate (give me more elephant examples). ElephantMind
ElephantMemory
ElephantRecall
TrunkTrack
PachydermMemory
MammothMemories
ElephantEcho
IvoryInsight
TuskerTips
ElephantWhisper
\item Prompt for "things that could go wrong" (the second part of vignette). Go through 40 examples and select ones, refine or ask for things long the same lines. 
\item Work for right format: length, examples, not giving the lesson at the end - easy
\item Inconsistency - need to start with a solid idea
\end{itemize} 


\section{Methodology 4: LLM Brainstorm}

\subsection{Methodology}
\begin{itemize}
\item Prompted with generate description + traditional designs. Added prompt for our complete sticky notes. 
\item Asked for help in generated new sticky notes - not useful
\item Asked it to come up with designs based on random combinations. Address lack of diversity
\item Example: Digital Manipulation, Genomics, Facial Recognition, Scary
\end{itemize}

\subsection{Designs}
Sample:

\design{Virtual Persona}{Virtual Persona is a social media platform that encourages users to create exaggerated online personas. Alice enjoys the freedom to express herself more boldly online but faces backlash when her boss finds her controversial posts, revealing the stark difference between acceptable behavior online and offline.}


\subsection{Discussion}



\section{Concluding Remarks}


